A White Noise Approach to Evolutionary Ecology

### 2020-04-06 .A White Noise Approach To Evolutionary Ecology/draft.Rmd
ALL of the rigorous proofs are finally complete

\begin{multline}\label{N_det}
\frac{d}{dt}N(t)=\frac{d}{dt}\int_{-\infty}^{+\infty}\nu(x,t)dx=\int_{-\infty}^{+\infty}\frac{\partial}{\partial t}\nu(x,t)dx \\
=\int_{-\infty}^{+\infty} m(\nu,x)\nu(x,t)dx+\int_{-\infty}^{+\infty}\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)dx \\
=N(t)\int_{-\infty}^{+\infty} m(\nu,x)p(x,t)dx=\bar m(t) N(t).
\end{multline}
Using our assumptions on mutation and rate of growth, we show in SM \S\ref{finite} $\nu(x,t)$ is twice differentiable with respect to $x$ and $\int_{-\infty}^{+\infty}\nu(x,t)dx<\infty$ for all $t\geq0$. This implies that we are justified in swapping the order of differentiation and integration and the result $\int_{-\infty}^{+\infty}\frac{\partial^2}{\partial x^2}\nu(x,t)dx=0$ can be derived from the fundamental theorem of calculus. Biological reasoning agrees with this latter result since mutation neither creates nor destroys individuals, but merely changes their type from their parental type. Taking the same approach, we derive the dynamics of the phenotypic distribution $p(x,t)$ in response to selection and muation as
\begin{multline}\label{p_PDE}
\frac{\partial}{\partial t}p(x,t)=\frac{\partial}{\partial t}\frac{\nu(x,t)}{N(t)}=\frac{1}{N^2(t)}\left(N(t)\frac{\partial}{\partial t}\nu(x,t)-\nu(x,t)\frac{d}{dt}N(t)\right) \\
=\frac{1}{N(t)}\Big(m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)-\bar m(t)\nu(x,t)\Big) \\
=(m(\nu,x)-\bar m(t)) p(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}p(x,t).
\end{multline}

This result closely resembles Kimura's continuum-of-alleles model [@Kimura1965; @Brger1986]. The primary difference being that our model utilizes diffusion instead of convolution with an arbitrary mutation kernel. Of course, our model of mutation can be derived as an approximation to Kimura's model, which has been referred to as the Gaussian allelic approximation in reference to the distribution of mutational effects at loci in a genome on the values of traits [@Lande1975; but see also @9780471986539], the infinitesimal model in reference to modelling continuous traits as being encoded by an infinite number of loci each having infinitesimal effect [@Barton2017] and the Gaussian descendants approximation in reference to offspring trait values being normally distributed around their parental values [@Turelli2017]. Alternatively, since diffusion is the continuous-time equivalent to convolution against a Gaussian kernel (SM \S\ref{diffconvequiv}), equation (\ref{p_PDE}) can also be seen as a special case of Kimura's continuum-of-alleles model. We return to these concepts in \S\ref{inheritance} where we review a simple model of inheritance that leads us to an expression for the stochastic evolution of additive genetic variance.

@@

To map the degree of niche overlap to fitness, we assume competition between individuals with niche locations $x_i$ and $x_j$ additively decreases the Malthusian fitness for the individual in species $i$ by $c_i\mathcal{O}_{ij}(x_i,x_j)$ for some $c_i\geq0$. We refer to $c_i$ as the sensitivity to competition for species $i$. The term $c_i\mathcal{O}_{ij}(x_i,x_j)$ coincides with a special case of a term used to capture competition in Dawson's geostochastic logistic model, an SPDE model developed to study the combined effects of demographic stochasticity, spatial dispersion and locally finite carrying capacity [@Dawson1978]. In SM \S\ref{diffuse} we combine our treatment of resource competition with equations (\ref{N}), (\ref{xbarfinal}) and (\ref{Gfinal}) to find

@@

Note that although $r_i(t)$ is referred to in the coexistence literature as the intrinsic growth rate of the population, $R_i$ is a deeper intrinsic quantity. For now we refer to $R_i$ as the _innate_ growth rate. Previous work has shown the importance of demographic stochasticity [@Schreiber2017] and evolutionary dynamics [@Schreiber2018] in determining ceoxistence of competing species. However, these studies consider evolution and stochasticity separately. With the connection outlined in this section formally established, researchers may pursue a postmodern coexistence theory that naturally includes both the evolutionary dynamics of populations and the effects of demographic stochasticity in a simple synthetic framework.

>>>>

\begin{multline}\label{N_det}
\frac{d}{dt}N(t)=\frac{d}{dt}\int_{-\infty}^{+\infty}\nu(x,t)dx=\int_{-\infty}^{+\infty}\frac{\partial}{\partial t}\nu(x,t)dx \\
=\int_{-\infty}^{+\infty} m(\nu,x)\nu(x,t)dx+\int_{-\infty}^{+\infty}\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)dx \\
=N(t)\int_{-\infty}^{+\infty} m(\nu,x)p(x,t)dx=\bar m(t) N(t).
\end{multline}
Using our assumptions on mutation and rate of growth, we show in SM \S\ref{finite} $\nu(x,t)$ is twice differentiable with respect to $x$ and $\int_{-\infty}^{+\infty}\nu(x,t)dx<\infty$ for all $t\geq0$. This implies that we are justified in swapping the order of differentiation and integration and the result $\int_{-\infty}^{+\infty}\frac{\partial^2}{\partial x^2}\nu(x,t)dx=0$ can be derived from the fundamental theorem of calculus. Biological reasoning agrees with this latter result since mutation neither creates nor destroys individuals, but merely changes their type from their parental type. Taking the same approach, we derive the dynamics of the phenotypic distribution $p(x,t)$ in response to selection and muation as

\begin{multline}\label{p_PDE}
\frac{\partial}{\partial t}p(x,t)=\frac{\partial}{\partial t}\frac{\nu(x,t)}{N(t)}=\frac{1}{N^2(t)}\left(N(t)\frac{\partial}{\partial t}\nu(x,t)-\nu(x,t)\frac{d}{dt}N(t)\right) \\
=\frac{1}{N(t)}\Big(m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)-\bar m(t)\nu(x,t)\Big) \\
=(m(\nu,x)-\bar m(t)) p(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}p(x,t).
\end{multline}
This result closely resembles Kimura's continuum-of-alleles model [@Kimura1965; @Brger1986]. The primary difference being that our model utilizes diffusion instead of convolution with an arbitrary mutation kernel. Of course, our model of mutation can be derived as an approximation to Kimura's model, which has been referred to as the Gaussian allelic approximation in reference to the distribution of mutational effects at loci in a genome on the values of traits [@Lande1975; but see also @9780471986539], the infinitesimal model in reference to modelling continuous traits as being encoded by an infinite number of loci each having infinitesimal effect [@Barton2017] and the Gaussian descendants approximation in reference to offspring trait values being normally distributed around their parental values [@Turelli2017]. Alternatively, since diffusion is the continuous-time equivalent to convolution against a Gaussian kernel (SM \S\ref{diffconvequiv}), equation (\ref{p_PDE}) can also be seen as a special case of Kimura's continuum-of-alleles model. We return to these concepts in \S\ref{inheritance} where we review a simple model of inheritance that leads us to an expression for the stochastic evolution of additive genetic variance.

@@

To map the degree of niche overlap to fitness, we assume competition between individuals with niche locations $x_i$ and $x_j$ additively decreases the Malthusian fitness for the individual in species $i$ by $c_i\mathcal{O}_{ij}(x_i,x_j)$ for some $c_i\geq0$. We refer to $c_i$ as the sensitivity to competition for species $i$. The term $c_i\mathcal{O}_{ij}(x_i,x_j)$ coincides with a special case of a term used to capture competition in Dawson's geostochastic logistic model, an SPDE model developed to study the combined effects of demographic stochasticity, spatial dispersion and locally finite carrying capacity [@Dawson1978]. In relation to the example fitness function discussed in \S\ref{deterministic}, consider $\kappa(x_i,x_j)=\mathcal{O}_{ij}(x_i,x_j)$. In SM \S\ref{diffuse} we combine our treatment of resource competition with equations (\ref{N}), (\ref{xbarfinal}) and (\ref{Gfinal}) to find

@@

Note that although $r_i(t)$ is referred to in the coexistence literature as the intrinsic growth rate of the population, $R_i$ is a deeper intrinsic quantity. For now we refer to $R_i$ as the _innate_ growth rate. Previous work has shown the importance of demographic stochasticity [@Schreiber2017] and evolutionary dynamics [@Schreiber2018] in determining ceoxistence of competing species. However, these studies consider evolution and demographic stochasticity separately. With the connection outlined in this section formally established, researchers may pursue a postmodern coexistence theory that naturally includes both the evolutionary dynamics of populations and the effects of demographic stochasticity in a simple synthetic framework.
